---
title: "Untitled"
author: "Ryan Harrington"
date: "5/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A detailed description of the CSV files uploaded to the drive and what they mean. 

`Student_Growth`: This file contains the average percent of student growth achieved in English/Language Arts, Mathematics, and English Language Proficiency. The statistics reported are those used within Delaware Student Success Framework (DSSF) since 2018. Each student has a growth target to achieve in ELA (grades 4-8), Math (grades 4-8), and, if an English Learner, in English Language Proficiency (grades K-12). The average percent of student growth achieved statistic shows, on average, how close a group of students came to achieving their target. (Source: Delaware Open Data)

`Student_Assessment_Performance`: This file contains the number of tested students, proficient students, and the corresponding proficiency rate for the key annual assessments administered during a given school year. (Source: Delaware Open Data)

`Unemployment_County`: Data on the unemployment rate and employment rate of the civilian labor force at the county level. This data is from 2010-2019. (Source: ACS)

`Unemployment_Census`: Data on the unemployment rate and employment rate of the civilian labor force at the census tract level. This data is from 2010-2019. (Source: ACS)

`Crime_Diseases`: Fatal crime rates and other health conditions like Arthritis, Asthma, Cancer, Cancer-related Deaths, Binge Drinking in the state of Delaware. This data is at the state level 2015 - 2020, couldn’t find it at a more granular level.


```{r}
library(tidyverse)
library(ggrepel)
library(ggtext)
library(here)
library(showtext)
library(tidycensus)

census_api_key(Sys.getenv("census_api_key"))

showtext_auto()

source(here::here("functions.R"))

theme_set(theme_minimal())
```

# Student Growth
Only 2 years of data: 2018 and 2019

```{r}
student_growth <- read_csv(here::here("Data", "Student_Growth.csv"))
```

```{r}
student_growth %>% 
  filter(District == "State of Delaware",
         Gender == "All Students",
         Race == "All Students",
         SpecialDemo == "All Students") %>% 
  ggplot(aes(x = SchoolYear,
             y = AvgPctOfTargetMet,
             color = Category)) +
  geom_line()
```

# Student Assessment

```{r}
student_assessment <- read_csv(here::here("Data", "Student_Assessment_Performance.csv"))
```

## 8th Grade Math

```{r}
student_assessment %>% 
  filter(District == "State of Delaware",
         Gender == "All Students",
         Race == "All Students",
         SpecialDemo == "All Students",
         ContentArea == "MATH",
         Grade == "8th Grade",
         `Assessment Name` == "Smarter Balanced Summative Assessment") %>% 
  ggplot(aes(x = `School Year`,
             y = PctProficient)) +
  geom_line()
```

```{r}
selected_districts <- 
  c("Brandywine School District", 
    "Christina School District", 
    "Colonial School District", 
    "Red Clay Consolidated School District", 
    "State of Delaware")
```

```{r}
create_student_proficiency_plot(data = student_assessment, 
                                content = "MATH", 
                                grade = "8th Grade", 
                                districts = selected_districts, 
                                title_copy = "Percentage of 8th Grade Students Proficient in Math by Year")
```

# 3rd Grade Reading

```{r}
student_assessment %>% 
  filter(District == "State of Delaware",
         Gender == "All Students",
         Race == "All Students",
         SpecialDemo == "All Students",
         ContentArea == "ELA",
         Grade == "3rd Grade",
         `Assessment Name` == "Smarter Balanced Summative Assessment") %>% 
  ggplot(aes(x = `School Year`,
             y = PctProficient)) +
  geom_line()

```

```{r}
create_student_proficiency_plot(data = student_assessment, 
                                content = "ELA", 
                                grade = "3rd Grade", 
                                districts = selected_districts, 
                                title_copy = "Percentage of 3rd Grade Students Proficient in ELA by Year")
```


# Crime Diseases

```{r}
crime_files <- list.files(here::here("Data", "Crime_Diseases"), full.names = TRUE)

crime_diseases <- 
  crime_files %>% 
  purrr::map_df(~ read_csv(.))
```

```{r}
crime_diseases %>% 
  filter(`State Name` == "Delaware") %>% 
  count(`Measure Name`, sort = T) %>% 
  filter(str_detect(`Measure Name`, "Racial"))
```

## Children in Poverty

```{r}
crime_diseases %>% 
  filter(`State Name` == "Delaware",
         `Measure Name` == "Children in Poverty",
         Source == "U.S. Census Bureau, American Community Survey") %>% 
  ggplot(aes(x = Edition,
             y = Value)) +
  geom_line()
```

## High School Graduation Rate

```{r}
americas_health_rankings_plot(
  data = crime_diseases,
  measure = "High School Graduation",
  title_text = "High School Graduation Rate in DE by Year",
  subtitle_text = "Percentage of high school students who graduated with a regular high school diploma within four years of starting ninth grade"
  ) +
  scale_y_continuous(labels = scales::label_percent(scale = 1),
                     limits = c(0, 100))
```

```{r}
americas_health_rankings_plot(
  data = crime_diseases,
  measure = "Violent Crime",
  title_text = "Violent Crime in DE per 100,000 People by Year",
  subtitle_text = "Number of murders, rapes, robberies and aggravated assaults per 100,000 population"
  )
```

# Census

## Function
```{r}
get_acs_with_year <- function(geo_var = "tract",
                              year_var = 2018,
                              var_nums = c("B19013_001"),
                              state_abbr = "DE",
                              survey_var = "acs5") {
  
  get_acs(geography = geo_var,
          variables = var_nums,
          state = state_abbr,
          geometry = FALSE,
          survey = survey_var,
          year = year_var,
          show_call = TRUE) %>% 
    mutate(year = year_var)
  
}
```

## Variables

```{r}
baselines <- c(
  population_count = "B01003_001",
  household_count  = "B25001_001"
)

housing_variables <- c(
  total_housing_units = "B25002_001",
  occupied_units      = "B25002_002",
  vacant_units        = "B25002_003",
  other_vacant_units  = "B25004_008"
  # B25002_002E/NULLIF(B25002_003E,0) as census_housing_Occupied_Vacant_Ratio,
  # census_housing_Total_Owned_Units = "B25106_002E",
  # census_housing_Total_Rented_Units = "B25106_024E",
  # B25106_002E/NULLIF(B25106_024E,0) as census_housing_Owned_Rented_Ratio, 
  # B25106_006E + B25106_010E + B25106_014E + B25106_018E + B25106_022E as census_housing_High_Living_Expenses_Units_Owned,
  # B25106_028E + B25106_032E + B25106_036E + B25106_040E + B25106_044E as census_housing_High_Living_Expenses_Units_Rented
)


health_variables <- c(
    health_total   = "B27001_001",
    medicare_total = "C27006_001"
    # B27001_004E + B27001_032E as census_health_Health_Coverage_Under_6,
    # B27001_007E + B27001_035E as census_health_Health_Coverage_6_18,
    # B27001_010E + B27001_038E as census_health_Health_Coverage_19_25,
    # B27001_013E + B27001_041E as census_health_Health_Coverage_26_34,
    # B27001_016E + B27001_044E as census_health_Health_Coverage_35_44,
    # B27001_019E + B27001_047E as census_health_Health_Coverage_45_54,
    # B27001_022E + B27001_050E as census_health_Health_Coverage_55_64,
    # B27001_025E + B27001_053E as census_health_Health_Coverage_65_74,
    # B27001_028E + B27001_056E as census_health_Health_Coverage_75_Over,
    # C27006_004E + C27006_014E as census_health_Medicare_Coverage_Under_19,
    # C27006_007E + C27006_017E as census_health_Medicare_Coverage_19_64,
    # C27006_010E + C27006_020E as census_health_Medicare_Coverage_65_Over
  )

economy_variables <- c(
  employed_population                             = "B23025_004",
  # income_social_security                          = "B19055_002",
  income_poverty_ratio_50_Below                   = "C17002_002",
  income_poverty_ratio_50_99                      = "C17002_003",
  income_poverty_ratio_100_124                    = "C17002_004",
  income_poverty_ratio_125_149                    = "C17002_005",
  income_poverty_ratio_150_184                    = "C17002_006",
  income_poverty_ratio_185_199                    = "C17002_007",
  income_poverty_ratio_200_Up                     = "C17002_008",
  total_households_received_food_stamps           = "B22007_002",
  # NULLIF(B19126_001E, -666666666) as census_economy_Median_Income,
  household_received_foodstamps_in_last_12_months = "B22002_002",
  number_unemployed                               = "B23025_005"
)

education_variables <- c(
  high_school_graduates = "B06009_003",
  bachelor_degrees      = "B06009_005"
)
```


## Wilmington

Source: https://www.wilmingtonde.gov/about-us/about-the-city-of-wilmington/population-demographics
Census tract map: https://opendata.firstmap.delaware.gov/datasets/delaware-census-tracts-2010/explore?location=39.711303%2C-75.541219%2C12.91

```{r}
census_tract_numbers <- c(
  # Boulevard
  "3", "4", "5",
  
  # Browntown / Hedgeville
  "25", "26", "27",
  
  # Central and Midtown Brandywine
  "28",
  
  # Delaware Avenue / Bancroft Parkway
  "11", "12", "13",
  
  # East Side
  "9", "29",
  
  # Northwest
  "2",
  
  # Price's Run
  "6.01", "6.02",
  
  # Riverside
  "30.02",
  
  # Southwest Wilmington
  "24",
  
  # South Wilmington
  "19.02",
  
  # West Center City
  "16", "21",
  
  # West Side
  "14", "15", "22", "23"
)
```


```{r}
census_tract_names <- glue::glue("Census Tract {census_tract_numbers}, New Castle County, Delaware")
```

```{r}
census_df_wilm <-
  2011L:2019L %>% 
  map_df(~ get_acs_with_year(geo_var = "tract",
                             year_var = .,
                             var_nums = c(baselines, housing_variables, economy_variables, education_variables)))
```

```{r}
clean_names <- 
  tribble(
    ~variable,                                                ~clean_name,
    "below_poverty_per_capita",                               "Below Poverty",
    "employed_population_per_capita",                         "Employed",
    "high_school_graduates_per_capita",                       "High School Graduates",
    "household_received_foodstamps_in_last_12_months_per_hh", "Received Foodstamps Last 12 Months",
    "total_households_received_food_stamps_per_hh",           "Received Foodstamps",
    "bachelor_degrees_per_capita",                            "Bachelor Degrees",
    "number_unemployed_per_capita",                           "Unemployed"
  )
```

```{r}
plot_df <- 
  census_df_wilm %>% 
  filter(NAME %in% census_tract_names) %>% 
  select(-moe) %>% 
  pivot_wider(names_from = "variable",
              values_from = "estimate") %>% 
  mutate(below_poverty = income_poverty_ratio_50_Below, income_poverty_ratio_50_99) %>% 
  mutate(across(c(high_school_graduates, bachelor_degrees, employed_population, 
                  number_unemployed, below_poverty), 
                ~ . / population_count, 
                .names = "{.col}_per_capita")) %>% 
  mutate(across(c(household_received_foodstamps_in_last_12_months, total_households_received_food_stamps), 
                ~ . / household_count, 
                .names = "{.col}_per_hh")) %>%
  pivot_longer(cols = c(-GEOID, -NAME, -year),
               names_to = "variable",
               values_to = "estimate") %>% 
  filter(str_detect(variable, "per_capita|per_hh")) %>% 
  filter(variable != "total_households_received_food_stamps_per_hh") %>% 
  group_by(year, variable) %>% 
  summarize(estimate = median(estimate)) %>% 
  ungroup() %>% 
  mutate(type = str_detect(variable, "per_capita"))

plot_df_last <-
  plot_df %>% 
  filter(year == max(year)) %>% 
  left_join(clean_names, by = "variable") %>% 
  mutate(clean_name = as.character(glue::glue("{scales::percent(estimate)} {clean_name}")))

ggplot() +
  geom_line(data = plot_df,
            aes(x = year,
                y = estimate,
                group = variable,
                color = type),
            size = 1,
            linejoin = "round",
            lineend = "round") +
  geom_point(data = plot_df_last,
             aes(x = year,
                 y = estimate,
                 group = variable,
                 color = type)) +
  scale_x_continuous(breaks = seq(2011, 2019, 2)) +
  scale_y_continuous(labels = scales::percent,
                     breaks = seq(0, 0.5, 0.25),
                     limits = c(0, 0.5),
                     sec.axis = dup_axis(
                       breaks = plot_df_last$estimate,
                       labels = plot_df_last$clean_name,
                       name = NULL)) +
  scale_color_manual(values = c(fse_blue(), fse_red())) +
  labs(x = "",
       y = "",
       title = "<b>Wilmington: Don't Stick to the Status Quo</b>",
       subtitle = "Percentage of key metrics for <span style = 'color: #d23742'><b>households</b></span> and <span style = 'color: #27829e'><b>people</b></span> by year in Wilmington, DE",
       caption = "Source: American Community Survey") +
  theme(legend.position = "none",
        plot.title = element_markdown(),
        plot.subtitle = element_markdown(),
        plot.title.position = "plot",
        panel.grid.minor = element_blank())
```

## State vs. Country

```{r}
census_df_state <-
  2011L:2019L %>% 
  map_df(~ get_acs_with_year(geo_var = "state",
                             year_var = .,
                             var_nums = c(baselines, housing_variables, economy_variables, education_variables)))

census_df_us <-
  2011L:2019L %>% 
  map_df(~ get_acs_with_year(geo_var = "us",
                             state_abbr = NULL,
                             year_var = .,
                             var_nums = c(baselines, housing_variables, economy_variables, education_variables)))
```

```{r}
census_df <-
  union(census_df_us, census_df_state)
```

```{r}
plot_df <- 
  census_df %>% 
  select(-moe) %>% 
  pivot_wider(names_from = "variable",
              values_from = "estimate") %>% 
  mutate(below_poverty = income_poverty_ratio_50_Below, income_poverty_ratio_50_99) %>% 
  mutate(across(c(high_school_graduates, bachelor_degrees, employed_population, 
                  number_unemployed, below_poverty), 
                ~ . / population_count, 
                .names = "{.col}_per_capita")) %>% 
  mutate(across(c(household_received_foodstamps_in_last_12_months, total_households_received_food_stamps), 
                ~ . / household_count, 
                .names = "{.col}_per_hh")) %>%
  pivot_longer(cols = c(-GEOID, -NAME, -year),
               names_to = "variable",
               values_to = "estimate") %>% 
  filter(str_detect(variable, "per_capita|per_hh")) %>% 
  filter(variable != "total_households_received_food_stamps_per_hh") # %>% 
  # group_by(year, NAME, variable) %>% 
  # summarize(estimate = median(estimate)) %>% 
  # ungroup() %>% 
  # mutate(type = str_detect(variable, "per_capita"))

plot_df_last <-
  plot_df %>% 
  filter(year == max(year)) %>% 
  left_join(clean_names, by = "variable") %>% 
  mutate(clean_name = as.character(glue::glue("{scales::percent(estimate)} {clean_name}")))

ggplot() +
  geom_line(data = plot_df,
            aes(x = year,
                y = estimate,
                group = NAME,
                color = NAME),
            size = 1,
            linejoin = "round",
            lineend = "round") +
  geom_point(data = plot_df_last,
             aes(x = year,
                 y = estimate,
                 group = NAME,
                 color = NAME)) +
  scale_x_continuous(breaks = seq(2011, 2019, 2)) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(values = c(fse_blue(), fse_red())) +
  facet_wrap(~variable, nrow = 2, scales = "free_y") +
  labs(x = "",
       y = "",
       title = "<b>Wilmington: Don't Stick to the Status Quo</b>",
       subtitle = "Percentage of key metrics for <span style = 'color: #d23742'><b>households</b></span> and <span style = 'color: #27829e'><b>people</b></span> by year in Wilmington, DE",
       caption = "Source: American Community Survey") +
  theme(legend.position = "none",
        plot.title = element_markdown(),
        plot.subtitle = element_markdown(),
        plot.title.position = "plot",
        panel.grid.minor = element_blank())
```